\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-21 paper.tex
save

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N$, simulator $g$;\\
        {\it Outputs:} proposal distribution $q_\psi(\theta)$;\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$. 
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \nabla_\psi \gamma q_\psi(\theta_m) \log q_\psi(\theta_m))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by evaluating the exact and known gradient of $U_d$ over
mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate with mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
In practice, using
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
\end{equation}
works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the adversarial variational
optimization procedure when coupled with WGAN-GP.

>>>>

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r$, simulator $g$.\\
        {\it Outputs:} proposal distribution $q_\psi(\theta)$, such that $p_r \approx p_\psi$.\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \gamma  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by evaluating the exact and known gradient of $U_d$ over
mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate with mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
In practice, using
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
\end{equation}
works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs. \glnote{add note on rmsprop vs. adam}
